Indicators for Ecosystem Condition in Norway

### 2023-10-04 NDVI_natopen.Rmd

## Introduction
The Normalized difference vegetation index (NDVI) can be used to describe the greenness of an area, and may thus represent a proxy for the amount of chlorophyll and plant productivity in the given area. Different ecosystem types naturally have greenness signals of different intensity but also with a certain variance. Deviations beyond the natural variation in either direction may indicate a reduction in ecological condition of a given ecosystem. E.g. lower than normal NDVI may indicate excessive disturbance, while higher than normal NDVI may indicate regrowth and transitions towards more densely vegetated systems. In recent assessments of ecological condition for forest and mountain ecosystems, two different NDVI-based indicators were developed and applied. For forests, NDVI-values across the country were compared to a spatial reference NDVI-signal from protected areas (Framstad et al. 2020). For mountains, the extent and consistence of the temporal change of NDVI-values over the last twenty years was assessed (Framstad et al 2021). In both cases, the NDVI-product from the MODIS satellite reflectance imagery was used based on available maps of spatial ecosystem extent in Norway (i.e. national maps for the forest and mountain ecosystems).

@@

## About the underlying data
In the NDVI project for natopen systems, we use two kinds of data (coming as four datasets) for building indicators for ecological condition:

@@

### Representativity in time and space
#### check here ####
For natopen systems, the NiN data contain 2881 natopen polygons of known area size across mainland Norway. However, NiN-mapping is not planned and performed in a spatially representative way and thus both the ecosystem occurrences and their spatial extent have to be treated as spatially biased. The 2881 natopen polygons are distributed across regions and major ecosystem types in the following way:

@@

### Temporal coverage
The currently available NiN data on natopen ecosystems span a time period from 2018-2021 and thus represent a contemporary sample of ecosystem occurrences. Note that none of the polygons in the NiN-data has been visited more than once during that period.

Available Sentinel-2 satellite NDVI data span the years 2015-2022.
Available MODIS satellite NDVI data span the years 2000-2022.
Available Landsat satellite NDVI data span the years 1984-2021.

## Collinearities with other indicators
Unlike wetlands and semi.natural systems, natopen areas are not subject to shrub and tree encroachment due to changes in land-use, Thus, we do not expect any co-linearity of the NDVI signal with e.g. the re-growth indicator based on LiDAr derived canopy height.

## Reference state and values
### Reference state and scaling values
This indicator project is mainly exploratory, mapping the possibilities of and lacking prerequisites for satellite derived NDVI data to inform ecological condition assessments. Therefore, we lack a concise strategy for how to define the reference state, especially prior to having access to an ecosystem map for natopen ecosystems (cf. Jakobsson et al. 2020, Töpper & Jakobsson 2021). In this exploratory work, we will use the condition classification in the NiN-data to explore (1) how NDVI connects to natopen ecology and condition and (2) its applicability for defining a reference state. We will use statistical models to test whether or not NDVI varies systematically with condition and across time.

## Uncertainties
Given a condition index can be achieved, we can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a measure of spatial uncertainty for a geographical area (see Töpper & Jakobsson 2021).

## References

@@

## Analyses
### Data sets
The analyses in this document make use of the following data sets:

@@

And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway
```{r message=FALSE}
# Import region- og Norgeskart, make sure they have the same geometry as the nin data
nor <- st_read("data/outlineOfNorway_EPSG25833.shp") %>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nin))

@@

# change region names to something R-friendly
reg$region
reg$region <- c("Northern Norway", "Central Norway", "Eastern Norway", "Western Norway", "Southern Norway")

@@

#### NDVI data
All NDVI data are calculated mean NDVI values for each respective NiN polygon from each available Sentinel-2, MODIS, or Landsat image in Google Earth Engine (GEE). 
```{r, echo = F}
# downloading the unprocessed NDVI data from a data-cache et the NINA server
df.s <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Sentinel.raw.RDS")
df.m <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Modis.raw.RDS")
df.l <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Landsat.raw.RDS")
```

@@

### Data handling
#### NiN data
```{r, echo = T, eval = F}
# fixing variable- and ecosystem-names with special characters
colnames(nin)
colnames(nin)[c(3, 8, 17, 26, 31, 33, 34)] <- c(
  "hovedoekosystem", "kartleggingsaar", "noyaktighet",
  "omraadenavn", "uk_naertruet", "uk_sentraloekosystemfunksjon",
  "uk_spesieltdaarligkartlagt"
)
unique(nin$hovedoekosystem)

@@

@@

#### NDVI data
```{r, results='hide', message=FALSE}
## Sentinel-2
# join nin.natopen & Sentinel NDVI data
SentinelNDVI.natopen <- full_join(nin.natopen, df.s, by = "id")
# summary(SentinelNDVI.natopen)
SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
  mutate(
    hovedoekosystem = as.factor(hovedoekosystem),
    hovedtype = as.factor(hovedtype),
    ninkartleggingsenheter = as.factor(ninkartleggingsenheter),
    lokalitetskvalitet = as.factor(lokalitetskvalitet),
    tilstand = as.factor(tilstand),
    area_meters = st_area(SentinelNDVI.natopen)
  )
# summary(SentinelNDVI.natopen)
# get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(!is.na(hovedtype))
SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(!is.na(mean))
# summary(SentinelNDVI.natopen)
# get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
dim(SentinelNDVI.natopen)
SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(as.numeric(area_meters) >= 100)
dim(SentinelNDVI.natopen)
# split date into year, month & day
SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
  dplyr::mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date)
  )
# add column for sub-ecosystem types
SentinelNDVI.natopen <- SentinelNDVI.natopen %>% mutate(
  subtype = substring(ninkartleggingsenheter, 4),
  subtype = str_remove(subtype, "-"),
  subtype = str_remove(subtype, "-")
)
# we are using max NDVI per year in every NiN polygon
SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))

summary(SentinelNDVI.natopen)

@@

# summary(ModisNDVI.natopen)
ModisNDVI.natopen <- ModisNDVI.natopen %>%
  mutate(
    hovedoekosystem = as.factor(hovedoekosystem),
    hovedtype = as.factor(hovedtype),
    ninkartleggingsenheter = as.factor(ninkartleggingsenheter),
    lokalitetskvalitet = as.factor(lokalitetskvalitet),
    tilstand = as.factor(tilstand),
    area_meters = st_area(ModisNDVI.natopen)
  )
# summary(ModisNDVI.natopen)
# get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(!is.na(hovedtype))
ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(!is.na(mean))
# summary(ModisNDVI.natopen)
# get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
dim(ModisNDVI.natopen)
ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(as.numeric(area_meters) >= 62500)
dim(ModisNDVI.natopen)
# split date into year, month & day
ModisNDVI.natopen <- ModisNDVI.natopen %>%
  dplyr::mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date)
  )
# we are using max NDVI per year in every NiN polygon
ModisNDVI.natopen <- ModisNDVI.natopen %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))
# summary(ModisNDVI.natopen)

@@

# summary(LandsatNDVI.natopen)
LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
  mutate(
    hovedoekosystem = as.factor(hovedoekosystem),
    hovedtype = as.factor(hovedtype),
    ninkartleggingsenheter = as.factor(ninkartleggingsenheter),
    lokalitetskvalitet = as.factor(lokalitetskvalitet),
    tilstand = as.factor(tilstand),
    area_meters = st_area(LandsatNDVI.natopen)
  )
# summary(LandsatNDVI.natopen)
# get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(!is.na(hovedtype))
LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(!is.na(mean))
# summary(LandsatNDVI.natopen)
# get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
dim(LandsatNDVI.natopen)
LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(as.numeric(area_meters) >= 900)
dim(LandsatNDVI.natopen)
# split date into year, month & day
LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
  dplyr::mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date)
  )
# we are using max NDVI per year in every NiN polygon
LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))

@@

### Exploratory analyses, Sentinel-2

**Where are the natopen sites?**
```{r}
tm_shape(regnor) +
  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE) +
  tm_borders() +
  tm_shape(SentinelNDVI.natopen) +
  tm_dots("hovedtype", midpoint = NA, palette = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE), scale = 2.5, legend.show = FALSE) +
  tm_add_legend(
    type = "symbol",
    col = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),
    labels = c(
      "T12 - coastal meadows",
      "T18 - flood plains",
      "T2 - shallow soils",
      "T20 - innfrysningsmark",
      "T21 - sand dunes"
    ),
    title = "Major ecosystem type",
    size = 3
  ) +
  tm_layout(
    main.title = "Naturally open ecosystems below the tree line (NiN, Miljødirektoratets instruks)",
    main.title.size = 1.2, legend.position = c("right", "bottom"),
    legend.text.size = 1.3, legend.title.size = 1.4
  )
```

**How are polygon sizes distributed?**
```{r}
summary(SentinelNDVI.natopen$area_meters)
hist(SentinelNDVI.natopen$area_meters, xlim = c(0, 20000), breaks = 50000)
abline(v = 100, lty = 2)
```

@@

**How does NDVI vary over the years?**
```{r}
SentinelNDVI.natopen %>%
  ggplot(aes(x = year, y = mean)) +
  geom_point() +
  facet_grid(tilstand ~ hovedtype)
```

@@

**How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)**
```{r warning=FALSE}
SentinelNDVI.natopen %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = hovedtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey")

SentinelNDVI.natopen$subtype <- factor(SentinelNDVI.natopen$subtype, levels = c("C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8", "E1"))
SentinelNDVI.natopen %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = subtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 6))
```

@@

**How does NDVI vary across regions in natopens (major types) in good condition? (using only NDVI data matching NiN-mapping years)**
```{r warning=FALSE}
SentinelNDVI.natopen %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x = region, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
```

@@

**How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)?**
```{r warning=FALSE}
SentinelNDVI.natopen %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

The plots suggest a slight decrease in NDVI as condition deteriorates. Overlaps between condition classes are generally very large.
#### continue here ####

@@

### Regression analyses Sentinel-2, NDVI as a function of condition
We can investigate further how NDVI and condition are connected using regression analyses.

First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
```{r warning=FALSE, message=FALSE}
# NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
SentinelNDVI.natopen %>%
  group_by(id, year) %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_grid(region ~ hovedtype)
```

T20 is has only very few data and is present in only 1 conditino class and 2 regions, we omit it.
Northern Norway does not have any T2 occurrences, central Norway doesn't have any T21 occurrences, we won't be able to model these types across all regions.

NDVI data are bounded between -1 and 1, and thus require modelling with an appropriate method that can handle bounded data. We can transform the variable to be bounded between 0 and 1 and use beta-regression models:
```{r}
SentinelNDVI.natopen$mean_beta <- (SentinelNDVI.natopen$mean + 1) / 2
# NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
# we drop T41 for the analysis as it lacks data for most combinations of condition and region, and thus would cause convergence issues
SentinelNDVI.natopen.train <- SentinelNDVI.natopen %>%
  filter(year == kartleggingsaar) %>%
  filter(hovedtype != "T20")

# check if there's any 0s or 1s (which beta cannot handle)
summary(SentinelNDVI.natopen.train$mean_beta) # ok

# We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
# we exclude Northern Norway and T21 for this first exploratory model
model.natopen.cond.Sent <- betareg(mean_beta ~ tilstand_num * region * hovedtype, data = SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region != "Northern Norway" & SentinelNDVI.natopen.train$hovedtype != "T21", ])
model.natopen.cond.Sent <- StepBeta(model.natopen.cond.Sent)
summary(model.natopen.cond.Sent)
```
NDVI values do vary between:
- Regions
- Ecosystem types
- Condition classes

It is evident that there are differences between regions. But since we are not interested in the statistics for region differences, we run separate models for every region for ease of interpretation.
```{r, include = F}
rm(model.natopen.cond.Sent)
```

@@

# Northern Norway
```{r message=FALSE, warning=FALSE}
SentinelNDVI.natopen.train %>%
  filter(region == "Northern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)

summary(model.natopen.cond.Sent.N)$coefficients
```
Not much of a pattern in either model or figure.

# Central Norway
```{r message=FALSE, warning=FALSE}
SentinelNDVI.natopen.train %>%
  filter(region == "Central Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)

summary(model.natopen.cond.Sent.C)$coefficients
```
NDVI in good condition: T12/T18 > T2

>>>>

## Introduction {#intro-ndvin}
The Normalized difference vegetation index (NDVI) can be used to describe the greenness of an area, and may thus represent a proxy for the amount of chlorophyll and plant productivity in the given area. Different ecosystem types naturally have greenness signals of different intensity but also with a certain variance. Deviations beyond the natural variation in either direction may indicate a reduction in ecological condition of a given ecosystem. E.g. lower than normal NDVI may indicate excessive disturbance, while higher than normal NDVI may indicate regrowth and transitions towards more densely vegetated systems. In recent assessments of ecological condition for forest and mountain ecosystems, two different NDVI-based indicators were developed and applied. For forests, NDVI-values across the country were compared to a spatial reference NDVI-signal from protected areas (Framstad et al. 2020). For mountains, the extent and consistence of the temporal change of NDVI-values over the last twenty years was assessed (Framstad et al 2021). In both cases, the NDVI-product from the MODIS satellite reflectance imagery was used based on available maps of spatial ecosystem extent in Norway (i.e. national maps for the forest and mountain ecosystems).

@@

## About the underlying data {#about-the-data-ndvin}
In the NDVI project for natopen systems, we use two kinds of data (coming as four datasets) for building indicators for ecological condition:

@@

### Representativity in time and space {#rep-in-space-and-time-ndvin}
<!-- #### check here #### -->
For natopen systems, the NiN data contain 2881 natopen polygons of known area size across mainland Norway. However, NiN-mapping is not planned and performed in a spatially representative way and thus both the ecosystem occurrences and their spatial extent have to be treated as spatially biased. The 2881 natopen polygons are distributed across regions and major ecosystem types in the following way:

@@

### Temporal coverage {#temp-coverage-ndvin}
The currently available NiN data on natopen ecosystems span a time period from 2018-2021 and thus represent a contemporary sample of ecosystem occurrences. Note that none of the polygons in the NiN-data has been visited more than once during that period.

- Available Sentinel-2 satellite NDVI data span the years 2015-2022.
- Available MODIS satellite NDVI data span the years 2000-2022.
- Available Landsat satellite NDVI data span the years 1984-2021.

## Collinearities with other indicators {#collinearity-ndvin}
Unlike wetlands and semi.natural systems, natopen areas are not subject to shrub and tree encroachment due to changes in land-use, Thus, we do not expect any co-linearity of the NDVI signal with e.g. [the re-growth indicator](https://ninanor.github.io/ecosystemCondition/gjengroing.html) based on LiDAr derived canopy height.

## Reference state and values {#ref-ndvin}
### Reference state and scaling values {#ref2-ndvin}
This indicator project is mainly exploratory, mapping the possibilities of and lacking prerequisites for satellite derived NDVI data to inform ecological condition assessments. Therefore, we lack a concise strategy for how to define the reference state, especially prior to having access to an ecosystem map for natopen ecosystems (cf. Jakobsson et al. 2020, Töpper & Jakobsson 2021). In this exploratory work, we will use the condition classification in the NiN-data to explore 
1. how NDVI connects to natopen ecology and condition and 
2. its applicability for defining a reference state.

We will use statistical models to test whether or not NDVI varies systematically with condition and across time.

## Uncertainties {#uncertainties-ndvin}
Given a condition index can be achieved, we can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a measure of spatial uncertainty for a geographical area (see Töpper & Jakobsson 2021).

## References {#references-ndvin}

@@

## Analyses {#analyses-ndvin}
### Data sets {#datasets-ndvin}
The analyses in this document make use of the following data sets:

@@

And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway
```{r, message=FALSE}
# Import region- og Norgeskart, make sure they have the same geometry as the nin data
nor <- st_read("data/outlineOfNorway_EPSG25833.shp") %>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nin))

@@

# change region names to something R-friendly
  # reg$region
reg$region <- c("Northern Norway", "Central Norway", "Eastern Norway", "Western Norway", "Southern Norway")

@@

#### NDVI data {#ndvi-data-ndvin}
All NDVI data are calculated mean NDVI values for each respective NiN polygon from each available Sentinel-2, MODIS, or Landsat image in Google Earth Engine (GEE). 
```{r, echo = F}
# downloading the unprocessed NDVI data from a data-cache et the NINA server
df.s <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Sentinel.raw.RDS")
df.m <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Modis.raw.RDS")
df.l <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Landsat.raw.RDS")
```

@@

### Data handling {#data-handling-ndvin}
#### NiN data
```{r, echo = T, eval = F}
# fixing variable- and ecosystem-names with special characters
colnames(nin)
colnames(nin)[c(3, 8, 17, 26, 31, 33, 34)] <- c(
  "hovedoekosystem", "kartleggingsaar", "noyaktighet",
  "omraadenavn", "uk_naertruet", "uk_sentraloekosystemfunksjon",
  "uk_spesieltdaarligkartlagt"
)
unique(nin$hovedoekosystem)

@@

```{r, include=F, eval=F}
saveRDS(nin.natopen, paste0(here::here(), "/data/cache/nin.natopen.RDS"))
saveRDS(nin.natopen2, paste0(here::here(), "/data/cache/nin.natopen2.RDS"))
```

@@

#### NDVI data {#ndvi-data-ndvin}
```{r, results='hide', message=FALSE}
## Sentinel-2
# join nin.natopen & Sentinel NDVI data
SentinelNDVI.natopen <- full_join(nin.natopen, df.s, by = "id")
# summary(SentinelNDVI.natopen)
SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
  mutate(
    hovedoekosystem = as.factor(hovedoekosystem),
    hovedtype = as.factor(hovedtype),
    ninkartleggingsenheter = as.factor(ninkartleggingsenheter),
    lokalitetskvalitet = as.factor(lokalitetskvalitet),
    tilstand = as.factor(tilstand),
    area_meters = st_area(SentinelNDVI.natopen)
  )
# summary(SentinelNDVI.natopen)
# get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(!is.na(hovedtype))
SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(!is.na(mean))
# summary(SentinelNDVI.natopen)
# get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
  # dim(SentinelNDVI.natopen)
SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(as.numeric(area_meters) >= 100)
  # dim(SentinelNDVI.natopen)
# split date into year, month & day
SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
  dplyr::mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date)
  )
# add column for sub-ecosystem types
SentinelNDVI.natopen <- SentinelNDVI.natopen %>% mutate(
  subtype = substring(ninkartleggingsenheter, 4),
  subtype = str_remove(subtype, "-"),
  subtype = str_remove(subtype, "-")
)
# we are using max NDVI per year in every NiN polygon
SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))

  # summary(SentinelNDVI.natopen)

@@

# summary(ModisNDVI.natopen)
ModisNDVI.natopen <- ModisNDVI.natopen %>%
  mutate(
    hovedoekosystem = as.factor(hovedoekosystem),
    hovedtype = as.factor(hovedtype),
    ninkartleggingsenheter = as.factor(ninkartleggingsenheter),
    lokalitetskvalitet = as.factor(lokalitetskvalitet),
    tilstand = as.factor(tilstand),
    area_meters = st_area(ModisNDVI.natopen)
  )
# summary(ModisNDVI.natopen)
# get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(!is.na(hovedtype))
ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(!is.na(mean))
# summary(ModisNDVI.natopen)
# get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
  # dim(ModisNDVI.natopen)
ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(as.numeric(area_meters) >= 62500)
  # dim(ModisNDVI.natopen)
# split date into year, month & day
ModisNDVI.natopen <- ModisNDVI.natopen %>%
  dplyr::mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date)
  )
# we are using max NDVI per year in every NiN polygon
ModisNDVI.natopen <- ModisNDVI.natopen %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))
# summary(ModisNDVI.natopen)

@@

# summary(LandsatNDVI.natopen)
LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
  mutate(
    hovedoekosystem = as.factor(hovedoekosystem),
    hovedtype = as.factor(hovedtype),
    ninkartleggingsenheter = as.factor(ninkartleggingsenheter),
    lokalitetskvalitet = as.factor(lokalitetskvalitet),
    tilstand = as.factor(tilstand),
    area_meters = st_area(LandsatNDVI.natopen)
  )
# summary(LandsatNDVI.natopen)
# get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(!is.na(hovedtype))
LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(!is.na(mean))
# summary(LandsatNDVI.natopen)
# get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
  # dim(LandsatNDVI.natopen)
LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(as.numeric(area_meters) >= 900)
  # dim(LandsatNDVI.natopen)
# split date into year, month & day
LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
  dplyr::mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date)
  )
# we are using max NDVI per year in every NiN polygon
LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))

@@

### Exploratory analyses, Sentinel-2 {#exporatory-analyses-ndvin}

**Where are the natopen sites?**
```{r where-are-natopen-sites, fig.cap="A map of Norway showing the location of the naturally open ecosystem sites. T12 = coastal meadows, T18 = flood plains, T2 = shallow soils, T20 = innfrysningsmark, T21 = sand dunes."}
tm_shape(regnor) +
  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE) +
  tm_borders() +
  tm_shape(SentinelNDVI.natopen) +
  tm_dots("hovedtype", midpoint = NA, palette = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE), scale = 2.5, legend.show = FALSE) +
  tm_add_legend(
    type = "symbol",
    col = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),
    labels = c(
      "T12",
      "T18",
      "T2",
      "T20",
      "T21"
    ),
    title = "Major ecosystem type",
    size = 3
  ) +
  tm_layout(
    main.title = "Naturally open ecosystems below the tree line (NiN, Miljødirektoratets instruks)",
    main.title.size = 1.2, legend.position = c("right", "bottom"),
    legend.text.size = 1.3, legend.title.size = 1.4
  )
```

**How are polygon sizes distributed?**
```{r histogram-sentinall-ndvin, fig.cap="Histogram of naturally open ecosystem polygon sizes in Norway."}
#summary(SentinelNDVI.natopen$area_meters)
hist(SentinelNDVI.natopen$area_meters, xlim = c(0, 20000), breaks = 50000)
abline(v = 100, lty = 2)
```

@@

**How does NDVI vary over the years?**
```{r NDVI-values-over-time-ndvin, fig.cap="A figure showing NDVI values for naturally open ecosystem over time for six different main nature types in varying degrees of condition.}
SentinelNDVI.natopen %>%
  ggplot(aes(x = year, y = mean)) +
  geom_point() +
  facet_grid(tilstand ~ hovedtype)
```

@@

**How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)**
```{r NDVI-major-ndvin, warning=FALSE, fig.cap="NDVI values in major natopen ecosystem types in good condition."}
SentinelNDVI.natopen %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = hovedtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey")
```
```{r}
SentinelNDVI.natopen$subtype <- factor(SentinelNDVI.natopen$subtype, levels = c("C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8", "E1"))
```

```{r NDVI-basic-ndvin, warning=FALSE, fig.cap="NDVI values in basic natopen ecosystem types in good condition.}
SentinelNDVI.natopen %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = subtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 6))
```

@@

**How does NDVI vary across regions in natopens (major types) in good condition? (using only NDVI data matching NiN-mapping years)**
```{r ndvi-regions-ndvin, warning=FALSE, fig.cap="DVI variation across regions in natopen ecosystems (major types) in good condition.}
SentinelNDVI.natopen %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x = region, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
```

@@

**How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)?**
```{r ndvi-condition-ndvin, warning=FALSE, fig.cap="NDVI values across natopen sites in different ecological condition.}
SentinelNDVI.natopen %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

The plots suggest a slight decrease in NDVI as condition deteriorates. 
Overlaps between condition classes are generally very large.

<!-- #### continue here #### -->

@@

### Regression analyses Sentinel-2, NDVI as a function of condition {#regressions-ndvin}
We can investigate further how NDVI and condition are connected using regression analyses.

First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
```{r balanced-data-ndvin, warning=FALSE, message=FALSE, fig.cap="NDVI values across a condition gradient, faceted by natopen ecosystem types in five regions in Norway. ", fig.height=12}
# NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
SentinelNDVI.natopen %>%
  group_by(id, year) %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))+
  facet_grid(region ~ hovedtype)
```

T20 is has only very few data and is present in only 1 condition class and 2 regions, we omit it.
Northern Norway does not have any T2 occurrences, 
central Norway doesn't have any T21 occurrences, 
we won't be able to model these types across all regions.

NDVI data are bounded between -1 and 1, and thus require modelling with an appropriate method that can handle bounded data. 
We can transform the variable to be bounded between 0 and 1 and use beta-regression models:
```{r, message=F}
SentinelNDVI.natopen$mean_beta <- (SentinelNDVI.natopen$mean + 1) / 2
# NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
# we drop T41 for the analysis as it lacks data for most combinations of condition and region, and thus would cause convergence issues
SentinelNDVI.natopen.train <- SentinelNDVI.natopen %>%
  filter(year == kartleggingsaar) %>%
  filter(hovedtype != "T20")

# check if there's any 0s or 1s (which beta cannot handle)
  # summary(SentinelNDVI.natopen.train$mean_beta) # ok

# We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
# we exclude Northern Norway and T21 for this first exploratory model
model.natopen.cond.Sent <- betareg(mean_beta ~ tilstand_num * region * hovedtype, data = SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region != "Northern Norway" & SentinelNDVI.natopen.train$hovedtype != "T21", ])
model.natopen.cond.Sent <- StepBeta(model.natopen.cond.Sent)
summary(model.natopen.cond.Sent)
```

NDVI values do vary between:

- Regions
- Ecosystem types
- Condition classes

It is evident that there are differences between regions. 
But since we are not interested in the statistics for region differences, we run **separate models for every region** for ease of interpretation.
```{r, include = F}
rm(model.natopen.cond.Sent)
```

@@

**Northern Norway**
```{r north-nvdi-vs-tilstand-ndvin, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for natopen ecosystems in Northern Norway."}
SentinelNDVI.natopen.train %>%
  filter(region == "Northern Norway") %>%
  ggplot(aes(x = tilstand, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype)
```

```{r}
summary(model.natopen.cond.Sent.N)$coefficients
```

Not much of a pattern in either model or figure.
